Using dependent types for writting correct by construction tactics

### 2014-12-13 ICFP15paper/content/introduction.tex
for Edwin : Compiling version (with Idris 0.9.15.1) for automatic reflection of terms

Proving that one term is equal to another within some theories is something quite common in formal certification. Of course, the two things are usually not syntactically equal --this case would be trivial--, but they are equal according to a set of properties. These properties are either obtained by implementation (the implementation of + for natural numbers satisfies the property of associativity for example), or by axioms if we are working within an axiomatised structure. 
These proofs are usually omitted "on the paper" for the ordinary every day maths, but they are required by proof assistants, no matter how obvious they are. Currently, in the Idris\footnote{Idris is a relatively new, purely functional, and dependently typed programming language with full dependent types, which also supports tactic based proofs  (Idris will be briefly defined the the [Section X] (at the end of the paper))} programming language, they are done by hand, and they are made of a potentially long sequence of rewriting, where each step of rewriting uses one of the available properties. 
These proofs are the every day routine when we are working in a dependently typed language, as Idris, and especially when we're intensively indexing types over values in order to capture certain logical properties.
The indices can be for example natural numbers, and in this case the lemmas that we have to prove about these indices are arithmetic statements which use the properties of natural numbers and their usual operations : symmetry and commutativity for +, existence of neutral element for +, etc.

\subsection{A motivating little example}
We introduce a little example which shows how these proofs appear when we index a type by values like natural numbers.
Let's define the types of $Bit$ and $Binary$ number, both indexed over the value they represent, expressed in base ten as a natural number.
\begin{code}[caption=Bit and binary number, captionpos=b, label=lst1:haskell2]
data Bit : Nat -> Type where
     b0 : Bit Z
     b1 : Bit (S Z)

@@

It will be useful to be able to pad a binary number expressed with $w$ bits to a binary number with $S\ w$ digits : 
\begin{code}[caption=Padding, captionpos=b, label=lst1:haskell2]
pad : Binary w n -> Binary (S w) n
pad zero = zero # b0
pad (num # x) = pad num # x
\end{code}
Of course, this operation preserves the value represented by the binary number : this function takes a $Binary\ w\ n$ and produces a $Binary\ (S\ w)\ n$, so there will be no need to prove the correctness of this function after its definition (the $value$ is still $n$).

@@

In fact, this function produces a dependent pair. The first component is the value $bX$ represented by the first bit produced (of type $Bit\ bX$), and the second component is itself another dependent pair. This second dependent pair is composed of the value $bY$ represented by the second bit produced, and of a triple containing the two bits produced and the proof that these two bits are effectively encoding the result of the addition we had to perform.
For example, with the line corresponding to the computation $1_2 + 0_2 + 0_2 = (01)_2$, the function produces the bit $b0$ (encoding $0_2$) and the bit $b1$ (encoding $1_2$), and a proof that $1 + 0 + 0 = 1 + (2*0)$.

@@

Now, we want to define a function adding two binary numbers and a bit (used as a carry). We decide that this addition will only work for two binary numbers expressed with the same number of bits, and we produce a result with one more bit, and this result should of course represent the value $c + x + y$ if $c$ is representing the value of the input bit, and if $x$ and $y$ are representing the values of the two binary numbers given in input. Thus, we're tempted to write :
\begin{code}[caption=Addition of two binary numbers, captionpos=b, label=lst1:haskell2]
adc : Binary w x -> Binary w y -> Bit c 
     -> Binary (S w) (c + x + y)
adc zero zero carry = zero # carry
adc (numx # bX) (numy # bY) carry
   = let (v1 ** (v2 ** (carry0, lsb, _))) = 
      addBit bX bY carry in
          adc numx numy carry0 # lsb
\end{code}
Unfortunately, the types are mismatching for the two cases expressed by this pattern matching. The result of the first line $adc\ zero\ zero\ carry$ is expected to have the type\[Binary\ 1\ (plus\ (plus\ c\ 0)\ 0)\] and we're providing a term of type \[Binary\ 1\ (plus\ c\ (plus\ 0\ (plus\ 0\ 0)))\] [I AM SURPRISED HERE BY THE TYPE OF THE PROVIDED TERM].
The problem is even worse for the second line where the result is expected to have type \[plus\ (plus\ c\ (plus\ bit\ (plus\ v\ (plus\ v\ 0))))\ (plus\ bit\ (plus\ v\ (plus\ v\ 0)))\] while we're trying to provide a term of type \[plus\ v2\ (plus\ (plus\ (plus\ v1\ v)\ v)\ (plus\ (plus (plus\ v1\ v)\ v)\ 0))\]
For this kind of situation, Idris provides the possibility to write a provisional definition --provisional definitions are described with the brief overview of the language in page XX-- with the syntax "?=". Using a provisional definitions here will make the definition being accepted, but we will have to prove afterwards that the term provided can be converted into a term which has the right type.
However, for both of these two patterns, it is possible to prove the equality between the expected and the provided types, and this is even stronger than we need. Using this equality of types it will be trivial to produce a term of the expected type by using the provided term and by simply rewriting its type with this proof. These proofs will be made of a series of rewriting, and each rewriting will use a property about natural numbers and the usual operations.
For example, the proof for the second pattern look like this :
\begin{code}[caption=Proof by hand of the correspondence between expected and provided types for the second line of the definition of adc, captionpos=b, label=lst1:haskell2]
adc_lemma_2 = proof {
    intros;
    rewrite sym (plusZeroRightNeutral x);
    [...]
    rewrite sym (plusAssociative (plus c (plus bit0 (plus v v))) bit1 (plus v1 v1));
    rewrite (plusAssociative c (plus bit0 (plus v v)) bit1);
    rewrite plusCommutative bit1 (plus v v);
    [...]
    rewrite (plusAssociative (plus (plus x v) v1) (plus x v) v1);
    trivial;
}
\end{code}

@@

\begin{code}[caption=Code we'd like to write for the addition of two binary numbers, captionpos=b, label=lst1:haskell2]
adc : Binary w x -> Binary w y -> Bit c 
     -> Binary (S w) (c + x + y)
adc zero zero carry = zero # carry 
        by CommutativeGroupSolver
adc (numx # bX) (numy # bY) carry
   = let (v1 ** (v2 ** (carry0, lsb, _))) = 
      addBit bX bY carry in
          adc numx numy carry0 # lsb 
        by CommutativeGroupSolver
\end{code}
We would like to let a solver for commutative groups to find the proof on its own, and thereby, there will be no extra lemmas to prove afterwards.
This is a something doable because algebraic structures like monoid, group, commutative group and ring have all a very nice property : every element in them has a canonical represent. Thus these kind of proofs can be automatically produced by a tactic by simply computing the normal forms for each side of the equality, and then comparing them with the syntactical equality (Leibniz equality).

@@

Algebraic solvers, like ring solvers, are already implemented for various proof assistants, including Coq and Agda. In this paper, we describe a certified implementation of an automatic solver for equalities on algebraic structures, for the Idris language. The interesting thing that we're going to present in this paper lies in the way we construct it, in a "correct by construction" style.

In this paper, we focus on some specific theories, which are the following algebraic structures : magmas, semi-group, monoid, group and Abelian group.
The user will just have to prove that his structure satisfies the properties of an algebraic structure, and he will be allowed to invoke the corresponding tactic.

>>>>

Proving that one term is equal to another within some theories is something quite common in formal certification. Of course, the two things are usually not syntactically equal --this case would be trivial--, but they are equal according to a set of properties. These properties are either obtained by implementation (the implementation of + for natural numbers satisfies the property of associativity for example), or by axioms if we are working within an axiomatised structure. 
These proofs are usually omitted "on the paper" for the ordinary every day maths, but they are required by proof assistants, no matter how obvious they are. Currently, in the Idris\footnote{Idris is a relatively new, purely functional, and dependently typed programming language with full dependent types, which also supports tactic based proofs  (Idris will be briefly defined the the [Section X] (at the end of the paper))} programming language, they are done by hand, and they are made of a potentially long sequence of rewriting, where each step of rewriting uses one of the available properties. 
These proofs are the every day routine when we are working in a dependently typed language, as Idris, and especially when we're intensively indexing types over values in order to capture certain logical properties.
For example, If a type $T$ is indexed over natural numbers, then the definition of a function which has co-domain $T$ will often require to prove that the index of the produced term is equal to the expected index. This proof will use properties of natural numbers and their usual operations : symmetry and commutativity for +, existence of neutral element for +, etc.

\subsection{A motivating little example}
We introduce a little example which shows how these lemmas appear when we index a type by values like natural numbers.
Let's define the types of $Bit$ and $Binary$ number, both indexed over the value they represent, expressed in base ten as a natural number.
\begin{code}[caption=Bit and binary number, captionpos=b, label=lst1:haskell2]
data Bit : Nat -> Type where
     b0 : Bit Z
     b1 : Bit (S Z)

@@

It will be useful to be able to pad a binary number expressed with $w$ bits to a binary number with $S\ w$ digits : 
\begin{code}[caption=Padding, captionpos=b, label=lst1:haskell2]
pad : Binary w n -> Binary (S w) n
pad zero = zero # b0
pad (num # x) = pad num # x
\end{code}
Of course, this operation preserves the value represented by the binary number : this function takes a $Binary\ w\ n$ and produces a $Binary\ (S\ w)\ n$, so there will be no need to prove the correctness of this function after its definition since the $value$ is still $n$.

@@

In fact, this function produces a dependent pair. The first component is the value $bX$ represented by the first bit produced (of type $Bit\ bX$), and the second component is itself another dependent pair. This second dependent pair is composed of the value $bY$ represented by the second bit produced, and of a triple containing finally the two bits produced and the proof that these two bits are effectively encoding the result of the addition we had to perform.
For example, with the line corresponding to the computation $1_2 + 0_2 + 0_2 = (01)_2$, the function produces the bit $b0$ (encoding $0_2$) and the bit $b1$ (encoding $1_2$), and a proof that $1 + 0 + 0 = 1 + (2*0)$.

@@

Now, we want to define a function adding two binary numbers and a bit (used as a carry). We decide that this addition will only work for two binary numbers expressed with the same number of bits, and we produce a result with one more bit, and this result should of course represent the value $c + x + y$ if $c$ is representing the value of the input bit, and if $x$ and $y$ are representing the values of the two binary numbers given in input. Thus, we're tempted to write :
\begin{code}[caption=Addition of two binary numbers, captionpos=b, label=lst1:haskell2]
adc : Binary w x -> Binary w y -> Bit c 
     -> Binary (S w) (c + x + y)
adc zero zero carry = zero # carry
adc (numx # bX) (numy # bY) carry
   = let (v1 ** (v2 ** (carry0, lsb, _))) = 
      addBit bX bY carry in
          adc numx numy carry0 # lsb
\end{code}
Unfortunately, the types are mismatching for the two cases expressed by this pattern matching. The result of the first line $adc\ zero\ zero\ carry$ is expected to have the type\[Binary\ 1\ (plus\ (plus\ c\ 0)\ 0)\] but we're providing a term of type \[Binary\ 1\ (plus\ c\ (plus\ 0\ (plus\ 0\ 0)))\] [I AM SURPRISED HERE BY THE TYPE OF THE PROVIDED TERM].
The problem is even worse for the second line where the result is expected to have type \[plus\ (plus\ c\ (plus\ bit\ (plus\ v\ (plus\ v\ 0))))\ (plus\ bit\ (plus\ v\ (plus\ v\ 0)))\] while we're trying to provide a term of type \[plus\ v2\ (plus\ (plus\ (plus\ v1\ v)\ v)\ (plus\ (plus (plus\ v1\ v)\ v)\ 0))\]
For this kind of situation, Idris provides the possibility to write a provisional definition\footnote{provisional definitions are described with the brief overview of the language in page XX} with the syntax "?=". Using a provisional definitions here will make the definition being accepted, but we will have to prove afterwards that in both patterns, the term provided can be converted into a term which has the right type.
However, for both of these two patterns, it is possible to prove the equality between the expected and the provided types, and this is even stronger than we need. These proofs will be made of a series of rewriting, and each rewriting will use a property about the operation $+$ on natural numbers.
For example, the proof for the second pattern look like this :
\begin{code}[caption=Proof by hand of the correspondence between expected and provided types for the second pattern of the definition of adc, captionpos=b, label=lst1:haskell2]
adc_lemma_2 = proof {
    intros;
    rewrite sym (plusZeroRightNeutral x);
    [...]
    rewrite sym (plusAssociative (plus c (plus bit0 (plus v v))) bit1 (plus v1 v1));
    rewrite (plusAssociative c (plus bit0 (plus v v)) bit1);
    rewrite plusCommutative bit1 (plus v v);
    [...]
    rewrite (plusAssociative (plus (plus x v) v1) (plus x v) v1);
    trivial;
}
\end{code}

@@

\begin{code}[caption=Code we'd like to write for the addition of two binary numbers, captionpos=b, label=lst1:haskell2]
adc : Binary w x -> Binary w y -> Bit c 
     -> Binary (S w) (c + x + y)
adc zero zero carry ?= zero # carry 
        by CommutativeGroupSolver
adc (numx # bX) (numy # bY) carry
   ?= let (v1 ** (v2 ** (carry0, lsb, _))) = 
      addBit bX bY carry in
          adc numx numy carry0 # lsb 
        by CommutativeGroupSolver
\end{code}
We would like to let a solver for commutative groups to find the proof on its own.
This is a something doable because algebraic structures like monoid, group, commutative group and ring have all a very nice property : every element in them has a canonical represent. Thus these kind of proofs can be automatically produced by a tactic by simply computing the normal forms for each side of the equality, and then comparing them with the syntactical equality (Leibniz equality).

@@

Algebraic solvers, like ring solvers, are already implemented for various proof assistants, including Coq and Agda. In this paper, we describe a certified implementation of an automatic solver for equalities on algebraic structures, for the Idris language. The interesting thing that we're going to present in this paper lies in the way we construct it, which is a "correct by construction" style, with the same spirit as the functions we've written above.

In this paper, we focus on some specific theories, which are the following algebraic structures : magmas, semi-group, monoid, group and Abelian group.
